---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MixedUpSlimSims

<!-- badges: start -->
<!-- badges: end -->

MixedUpSlimSims is an R package to facilitate simulations
of admixed populations using SLiM, and particularly to collect the results
in terms of the ancestry tracts within the simulated individuals by way of
[tskit](https://tskit.dev/).


## Installation

You can install the development version of MixedUpSlimSims from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("eriqande/MixedUpSlimSims")
```

However, in order to use it, you will also have to set up some conda environments,
and, if you are working on Mac with Apple Silicon (M1 chip or higher) you actually
have to run your R/RStudio session using an x86_64 version of R that will
run through Rosetta on you Mac.

In order to get this to work on an a mac running Apple Silicon
there are a few things that must be done.

1.  Install a pyslim conda environment within the x86_64 shell.

    ```{sh}
    #| eval: false
    mamba create -n pyslim -c conda-forge pyslim
    ```

    Doing this requires that you have set up conda to work with x86_64. You can find
    instructions on how to do that [here](https://towardsdatascience.com/how-to-install-miniconda-x86-64-apple-m1-side-by-side-on-mac-book-m1-a476936bfaf0) or
    [here](https://taylorreiter.github.io/2022-04-05-Managing-multiple-architecture-specific-installations-of-conda-on-apple-M1/).
    Sadly, pyslim does not install with conda on arm64 because `kastore` is not compiled for that.  After that, activate the environment and do `pip install tspop`./
    
    Installing pyslim will get you `tskit` and any other conda packages that you need,
    except for a few others that we have added in there later.  Those additional
    packages must be installed into your pyslim environment.  You could do that
    when creating it, or be like us and just add them later:
    
    ```{sh}
    #| eval: false
    conda activate pyslim
    mamba install -c conda-forge -c bioconda bcftools bedtools
    ```

2.  Then, you need to have an x86_64 version of R that can run under RStudio. I recommend that you install [`rig`](https://github.com/r-lib/rig) to manage your R insreetallations. Once you have it installed, you can download and install the latest x86_64 version of R like this:

    ```{sh}
    #| eval: false
    rig add -a x86_64 release
    ```

    Then you can make sure that it was installed, and see what its name is by using `rig list`:

    ```{sh}
    #| eval: false
    rig list
        
    # the output should look something like this:
    * name        version    aliases
    ------------------------------------------
      4.2-arm64   (R 4.2.0)
    * 4.3-arm64   (R 4.3.3)
      4.3-x86_64  (R 4.3.3)  release
    ```

    And from that you can open this RStudio project by navigating to the project directory on your shell and doing:

    ```{sh}
    #| eval: false
    rig rstudio 4.3-x86_64 MixedUpSlimSims.Rproj
    ```

3.  Define the path to your pyslim conda environment, here. On my laptop it is:

    ```{r}
    PYSLIM_CONDA <- normalizePath("~/mambaforge-x86_64/envs/pyslim")
    ```
    
    But it probably will be different on yours.

