---
title: "Example Use of MixedUpSlimSims"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-use-of-MixedUpSlimSims}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks through a complete MixedUpSlimSims workflow.  We simulate
from a demographic model with asymmetric migration between two populations, using
allele frequencies similar to Westslope cutthroat trout (WCT) and rainbow trout
(RBT) in Cyclone Creek.  We then process the resulting SLiM tree sequence to
recover ancestry tracts, simulate diagnostic markers from those tracts, parse
the VCFs written by SLiM, and assemble data sets that can be used for downstream
method testing.

The main ingredients you need to bring to this workflow are:

- allele frequencies or founder genotypes for the source populations;
- chromosome or genome position information for those markers;
- a SLiM script that reads the founder genotypes, records tree sequences, and
  writes the expected output files; and
- a decision about which populations and years should be sampled and summarized.

The example below spells those steps out one at a time.  The
`slim_sim_a_dataset()` helper combines many of these steps when your SLiM script
and input files follow the package conventions.

## Initial set up

See the package README about configuring your system to be able to run this.
The main issue is to define the `MUP_CONDA` path as a system (Unix/Linux)
environment variable and then tell `reticulate` to use that environment before
Python is initialized.

After that you can get started.

```{r , message=FALSE}
library(reticulate)
library(tidyverse)
library(MixedUpSlimSims)
library(CKMRpop)

mup_conda <- Sys.getenv("MUP_CONDA")
if (mup_conda == "") {
  stop("Set MUP_CONDA to the full path of your MixedUpSlimSims conda environment.")
}

reticulate::use_condaenv(mup_conda, required = TRUE)
Sys.setenv(PATH = paste(file.path(mup_conda, "bin"), Sys.getenv("PATH"), sep = ":"))
```





## Making allele frequencies

This example starts from allele frequencies estimated for Cyclone Creek.  The
input file has one row per marker, with marker positions and population-specific
allele frequencies.
```{r}
cc_freqs <- read_csv("inputs/cyclone-creek-variable-markers-freqs.csv")
head(cc_freqs)
```

In this file, column `A` is WCT and column `B` is RBT.  In the SLiM model used
below, population 1 is the WCT-like population receiving some RBT ancestry, and
population 2 is the RBT-like population.  Therefore, we rename the frequency
columns to `af1` and `af2` in that order.

MixedUpSlimSims also needs each marker's cumulative genomic position: its
coordinate across the whole genome, not only within a chromosome.  The
`cumulative_genome_position()` helper adds those coordinates from chromosome
metadata.
```{r}
cc_freqs2 <- cc_freqs %>%
  mutate(chrom = as.character(chrom_f)) %>%
  select(-chrom_f) %>%
  cumulative_genome_position(mykiss_chroms) 

frqs <- cc_freqs2 %>%
  rename(af1 = A, af2 = B, POS = cPOS) %>%
  select(POS, af1, af2)
```



The SLiM script will initialize the first generation from a VCF named
`slim_input.vcf`.  Here we simulate 1000 diploid founder genotypes from each
source population and write them in the format that the example SLiM model
expects.
```{r}
simulate_founder_genos(frqs, N = c(1000, 1000), pop_names = c("wct", "rbt")) %>%
  write_input_vcf(out_path = "slim_input.vcf") 
```


## Running SLiM


The SLiM script controls the demography, migration, tree-sequence recording, and
which individuals are written to VCF.  Several example scripts are included in
`inst/SLiM-models`; this vignette uses the two-population, 10-generation example.
```{r}
# get the SLiM file.  It ships with the R package.
file <- system.file("SLiM-models/2-pop-10-gens-vcf-initialize.slim", package = "MixedUpSlimSims")
```

The full script is available on GitHub:
[2-pop-10-gens-vcf-initialize.slim](https://github.com/eriqande/MixedUpSlimSims/blob/main/inst/SLiM-models/2-pop-10-gens-vcf-initialize.slim).
For compatibility with the functions used below, the script needs to:

- read founder genotypes from `slim_input.vcf`;
- call `initializeTreeSeq()` and write a tree sequence named `SLiM.trees`;
- remember the individuals needed for ancestry tracing, especially the founder
  individuals whose population origins will define ancestry; and
- write sampled VCFs with names like `p1-10.vcf`, where the population and SLiM
  year can be parsed from the file name.

Now run the model with SLiM:
```{r}
# run it via system
system(paste("slim -s 333 ", file, collapse = " "))
```

This writes sampled VCF files for the years requested in the SLiM script:
```{r}
dir(pattern = "p[0-9]+-[0-9]+\\.vcf")
```

It also writes `SLiM.trees`, which records the ancestry information that `tskit`
will process.

## Processing the trees

SLiM time runs forward, while `tskit` time is measured backward from the final
generation.  In this example, SLiM year 11 is `tskit` time 0, SLiM year 10 is
`tskit` time 1, and so on back to SLiM year 1 at `tskit` time 10.  Since there
are 11 years of simulation, we retain nodes from `tskit` times 0 through 10 in
populations 1 and 2.
```{r}
# read the trees. Dump just captures the python code that was
# evaluated to get things in the py variable in the global env.
dump <- read_and_filter_trees(
  trees_path = "SLiM.trees",
  years_list = list(`1` = 0:10, `2` = 0:10)
)

```


`read_and_filter_trees()` loads and simplifies the tree sequence, leaving the
simplified `tskit` tree sequence in `py$ts`.  The next step turns the Python
nodes and individuals into ordinary R tibbles.
```{r}
ni <- ts_nodes_and_inds(py$ts)
```

The founder nodes define the population of origin for ancestry tracts.  In this
example they are the nodes that existed at `tskit` time 10, which corresponds to
SLiM year 1.
```{r}
founders <- founder_node_ids_and_pops(ni, 10)
```

The focal nodes are the individuals whose ancestry we want to summarize.  These
could be only the final sampled generations.  Here, to show the ancestry process
through time, we use all individuals from SLiM year 2 forward, which is `tskit`
time 9 down to 0, in both populations.

Population labels coming out of this SLiM simulation are base-1 indexed, so the
populations are named 1 and 2 rather than 0 and 1.
```{r}
focal_tibble <- tibble(
  pop = rep(c(1,2), each = 10),
  time = rep(0:9, 2)
)

focal_tibble
```

`get_focal_nodes()` converts those population-year choices into the node IDs
needed by the ancestry-linking code.
```{r}
fnl <- get_focal_nodes(
  ni, 
  Focal = focal_tibble
)
```

The object `fnl` keeps focal nodes divided by year.  `ancestral_segs()` then
cycles over those years and links each cohort back to the founders separately.
This avoids a problem in which older focal samples can become the linked
ancestors of younger samples, excluding the actual founders whose populations of
origin we need.

Now recover the ancestry segments for all focal individuals.
```{r}
segments_tib <- ancestral_segs(
  ts = py$ts,
  focal_nodes_list = fnl,
  founder_nodes = founders$nodes,
  nodes_tib = ni$nodes_tib,
  indiv_tib = ni$indiv_tib
)
```


## Plotting the _haploid_ ancestry segments

This is done with a simple function:
```{r, fig.height=10, fig.width=8}
plot_ancestry_of_genomes(segments_tib)
```


## Obtaining and plotting the ancestry tracts within each individual

The previous plot shows haploid ancestry segments for individual genomes.
`indiv_ancestry_tracts()` combines the two haplotypes within each diploid
individual, producing intervals labeled by diploid ancestry state.
```{r, message=FALSE}
indiv_segs <- indiv_ancestry_tracts(segments_tib)
```

Plot the diploid ancestry tracts:
```{r, fig.height=10, fig.width=8}
plot_ancestry_tracts(indiv_segs, chrom_ends = mykiss_chroms)
```

The left column is population 1 which is the WCT pop here. And the right column
is population 2 which is RBT.  

The colors denote the different ancestry combinations, which are coded in decimal
equivalents of trits or ternary numbers.  2 = 2 doses from p1; 4 = 1 dose from p1
and one from p2; 6 is two doses from p2.  This can extend to more than two
species.  

### How diploid ancestry states are encoded

The ancestry state is encoded as a ternary number, where each digit is the number
of copies inherited from one source population.  Each digit can be 0, 1, or 2.
For diploids, the digits must sum to 2.  The encoded value is then stored as the
decimal equivalent of that ternary number.  For example, in a three-species case:

- 002: two copies from species 1 --- $3^2 \times 0 + 3^1 \times 0 + 3^0 \times 2 = 2$
- 011: one copy from species 1 and one copy from species 2 --- $3^2 \times 0 + 3^1 \times 1 + 3^0 \times 1 = 4$
- 020: two copies from species 2 --- $3^2 \times 0 + 3^1 \times 2 + 3^0 \times 0 = 6$
- 200: two copies from species 3 --- $3^2 \times 2 + 3^1 \times 0 + 3^0 \times 0 = 18$

This encoding also extends naturally to simulations with more than two source
populations.

## Calculating admixture fractions from the ancestry tracts

Once the diploid ancestry tracts have been assembled, genome-wide admixture
fractions are a direct summary of those intervals.
```{r}
Q_values <- admixture_fracts_from_ancestry_tracts(indiv_segs)
```


## Getting into the VCFs

### A note on nomenclature

Throughout this package, `ped_id` means the pedigree ID of the individual.
SLiM's VCF output also has a `slimGenomePedigreeIDs` field, but those are the
pedigree IDs of the haplotypes or gametes, not of the diploid individuals.  In
the tibbles produced here, `ped_id`, `ped_p1`, and `ped_p2` refer to individual
pedigree IDs.

The function `slim_vcf2tib()` reads a single VCF into a long format. It uses
`bcftools` through a pipe.  Long format is not especially space-efficient, but it
is convenient for the moderate-size examples used here and for downstream
tidyverse operations.


```{r}
p1_10_genos <- slim_vcf2tib("p1-10.vcf")
head(p1_10_genos, n = 10)
```

Passing `long = FALSE` returns a wide tibble instead.
```{r}
p1_10_wide <- slim_vcf2tib("p1-10.vcf", long = FALSE)

# look at a piece of that
head(p1_10_wide[, 1:9], n = 10)
```

Most workflows read multiple VCF files.  The `slim_vcfs2tib()` helper lets you
specify the populations and SLiM years to read.  It parses the file names and
drops columns that are not usually needed downstream.  For example, to read SLiM
years 9, 10, and 11 from populations p1 and p2:
```{r}
sample_genos_long <- slim_vcfs2tib(Flist = list(p1 = 9:11, p2 = 9:11))
```

That produces a large tibble, but it makes later filtering and summarization
straightforward.

## Finding pairwise relationships among the sampled individuals

Because the SLiM model kept pedigree information, we can recover the known
relationships among sampled individuals.  This is useful for downstream tests of
parentage or close-kin methods, where, for example, one might want to know how
often aunts, uncles, or siblings are confused with parents.  The package wraps
the relationship-finding code in `pairwise_relationships()`.
```{r}
pairwise_relats <- pairwise_relationships(
  IT = ni$indiv_tib,
  sample_years = 0:2,
  num_gen = 2
)
```

Count the relationship classes found among the sampled individuals:
```{r}
pairwise_relats %>%
  count(dom_relat, max_hit)
```

Key to these relationships:

- `A` is "avuncular", meaning aunt-niece or uncle-niece, etc.
- `FC` is first cousin
- `GP` is grandparent-grandchild
- `PO1` is parent-offspring
- `Si` is sibling

The `max_hit` column records how many ancestors are shared between these individuals
at this level of relationship.  So `Si` with `max_hit = 1` is a half-sibling, while
`Si` with `max_hit = 2` is a full-sibling.

This simple Wright-Fisher example produces relatively few full siblings compared
with half siblings.  For analyses that need more full-sibling structure, a
non-Wright-Fisher SLiM model can be used instead.


## Simulating the diagnostic markers

The true ancestry tracts can also be used to simulate diagnostic markers.  The
example diagnostic marker set is included as package data.  Here we place those
markers into individuals from the last three years of the simulation.
```{r}
sim_diag <- example_spp_diag_markers %>%
  cumulative_genome_position(mykiss_chroms) %>%
  rename(diag_spp_idx = diag_spp) %>%
  place_diagnostic_markers(IndivSegs = indiv_segs %>% filter(ind_time %in% 0:2))
```

## Writing out those two data sets

The final step prepares the simulated marker data and the corresponding truth
data for downstream analyses.  We subsample individuals, format the diagnostic
markers, convert the variable-marker VCF data back to chromosome-specific
positions, and keep the true metadata, ancestry fractions, and ancestry tracts.
```{r}
set.seed(1)

subsamp <- sample(unique(sim_diag$ped_id), size = 2500, replace = FALSE)

sim_diag_ready <- sim_diag %>%
  filter(ped_id %in% subsamp) %>%
  rename(diag_spp = diag_spp_idx) %>%
  mutate(diag_spp = case_match(diag_spp, 1 ~ "WCT", 2 ~ "RBT", 3 ~ "YCT")) %>%
  rename(n = geno, indiv = ped_id) %>%
  mutate(indiv = as.character(indiv), pos = as.integer(pos), n = as.integer(n))

# Convert the variable markers back to their original chromosome-specific
# genome positions.
cc_freqs3 <- cc_freqs2 %>%
  select(pos, chrom, cPOS)

sim_var_snps_ready <- sample_genos_long %>%
  filter(ped_id %in% subsamp) %>%
  select(pos, ped_id, haplo, allele)  %>%
  rename(cPOS = pos) %>%
  left_join(cc_freqs3, by = join_by(cPOS)) %>%
  select(chrom, pos, ped_id, allele) %>%
  group_by(chrom, pos, ped_id) %>%
  summarise(n = sum(as.integer(allele))) %>%
  ungroup() %>%
  arrange(ped_id, chrom, pos) %>%
  rename(indiv = ped_id) %>%
  mutate(indiv = as.character(indiv), pos = as.integer(pos))

```

The object `Q_values` contains individual metadata, pedigree information, and
true admixture fractions.  We also keep the true ancestry segments for the same
individuals included in the marker data.
```{r}
Q_values

true_segs <- indiv_segs %>%
  filter(ped_id %in% subsamp)
```

These objects can be written together as one R object containing the simulated
observed data and the known truth:
```r
dir.create("results/004", showWarnings = FALSE, recursive = TRUE)
sim_results <- list(
  spp_diag = sim_diag_ready,
  var_snps = sim_var_snps_ready,
  metaQ = Q_values,
  true_segs = true_segs,
  cumul_genome = mykiss_chroms,
  alle_freqs = cc_freqs
)
write_rds(sim_results, file = "results/004/sim_results.rds", compress = "xz")
```
