---
title: "Example Use of MixedUpSlimSims"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-use-of-MixedUpSlimSims}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here we simulate from a demographic model that has asymmetric migration
between two species/populations and we provide allele frequencies like
those in Westslope cutthroat trout (WCT) and rainbow trout (RBT) in Cyclone Creek.

We also simulate the diagnostic markers given the ancestry
tracts of individuals.

By the end of it we will have a data set
with allele frequencies that is similar to Cyclone Creek and that should let us
start really testing out our methods.  

## Initial set up

See the package README about configuring your system to be able to run this.
The main issue is to define the `MUP_CONDA` path as both an R option and a
system (Unix/Linux) environment variable.  Note that this is done in the code
here since Rmarkdown seems to not read the `.Rprofile`.

After that you can get started.

```{r , message=FALSE}
library(reticulate)
library(tidyverse)
library(MixedUpSlimSims)
library(CKMRpop)  # this is necessary to use the MixedUpSlimSims::pairwise_relationship()
                  # until I fix how data get loaded in CKMRpop.

# The user should set the value of MUP_CONDA to point to the conda
# environment with all the things needed.
options(MUP_CONDA = "/Users/eriq/miniforge3/envs/mixed-up-parents")
Sys.setenv(MUP_CONDA = "/Users/eriq/miniforge3/envs/mixed-up-parents")


# set paths for python within R
reticulate::use_condaenv(condaenv = getOption("MUP_CONDA"))


# also set the path for system calls here
Sys.setenv(PATH = paste0(getOption("MUP_CONDA"), "/bin:", Sys.getenv("PATH")))
```





## Making allele frequencies

I am going to use the allele frequencies we estimated for Cyclone Creek.  I have
put those into `vignettes/inputs`.
```{r}
cc_freqs <- read_csv("inputs/cyclone-creek-variable-markers-freqs.csv")
head(cc_freqs)
```
A here is WCT and B is RBT. And, in our simulations, population 1 is like WCT getting a steady
flow of RBT and population 2 is like RBT just being RBT and continually coming into the WCT
habitat. So, we want A to be 1 and B to be 2.

Now we need to get them in the form that we need for `simulate_founder_genos()`.
We have to give the markers cumulative positions, meaning their position in the
total genome (not just within each chromosome).  We have a function for that now, too!
```{r}
cc_freqs2 <- cc_freqs %>%
  mutate(chrom = as.character(chrom_f)) %>%
  select(-chrom_f) %>%
  cumulative_genome_position(mykiss_chroms) 

frqs <- cc_freqs2 %>%
  rename(af1 = A, af2 = B, POS = cPOS) %>%
  select(POS, af1, af2)
```



We must write out an input file of founder genotypes
to `slim_input.vcf`.  So, we simulate 1000 indivs from each population and then write
those genotypes out.
```{r}
simulate_founder_genos(frqs, N = c(1000, 1000), pop_names = c("wct", "rbt")) %>%
  write_input_vcf(out_path = "slim_input.vcf") 
```


## Running SLiM


We can just run SLiM from here. We will run this in the current directory here.
We will use a system call, (though in Rmarkdown we could
have run it in a shell block).
```{r}
# get the SLiM file.  It ships with the R package.
file <- system.file("SLiM-models/2-pop-10-gens-vcf-initialize.slim", package = "MixedUpSlimSims")
```

Here is what the SLiM file look like:
```cpp

initialize() {
	defineConstant("n_p1", 1000); // number of individuals in the RBT pop
	defineConstant("n_p2", 1000); // number of individuals in the WCT pop

	initializeSLiMOptions(keepPedigrees=T); // turn on pedigree tracking; see section 25 of the manual (25.7, pg. 679)
	defineConstant("L", 1949962539); // length of genome
	initializeTreeSeq();
	initializeMutationRate(0); // zero mutations forward-in-time; will be added by msprime afterwards
	initializeMutationType("m1", 0.5, "f", 0.0); // neutral mutations
	m1.convertToSubstitution = F; // keep loci in the simulation, even if they reach fixation
	initializeGenomicElementType("g1", m1, 1.0);
	initializeGenomicElement(g1, 0, L);

	 // set up 29 trout chromosomes with not much recombination within them
    ends = c(84884017, 84884018, 170364868, 170364869, 255302337, 255302338, 340358758, 340358759, 432561311, 432561312, 515492034, 515492035, 595255810, 595255811, 679034094, 679034095, 747501830, 747501831, 818558021, 818558022, 898836325, 898836326, 988491333, 988491334, 1054543576, 1054543577, 1134902301, 1134902302, 1198270468, 1198270469, 1269166547, 1269166548, 1345694384, 1345694385, 1407413604, 1407413605, 1466989977, 1466989978, 1508401989, 1508401990, 1560331576, 1560331577, 1608881719, 1608881720, 1657923568, 1657923569, 1698286047, 1698286048, 1780887703, 1780887704, 1821070223, 1821070224, 1866387099, 1866387100, 1907331003, 1907331004, 1949962539);
    rates = c(1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09, 0.5, 1e-09);
    initializeRecombinationRate(rates, ends);
    initializeSex("A"); //Make all chromosomes autosomal
}
1 late() {
	//make two pops, with ongoing gene flow until the end of the simulation
	sim.addSubpop("p1", n_p1); // rainbow trout population
	sim.addSubpop("p2", n_p2); // westslope population

	// Read in a .vcf that was made in R and has p1 indivs first and p2 indivs second
	c(p1,p2).genomes.readFromVCF("slim_input.vcf", m1);


	sim.treeSeqRememberIndividuals(sim.subpopulations.individuals); // retains first-gen p1 and p2 individuals to determine genomic region population-of-origin;
	                                                                // automatically recorded by default, but doesn't hurt to keep explicitly
	p1.setMigrationRates(p2, 0.05); // migration rate from WCT into RBT
	p2.setMigrationRates(p1, 0.01); // migration rate from RBT into WCT
}


2: late() {
  // remember individuals over all generations so we can plot them and see how
  // the admixture is proceeding
  sim.treeSeqRememberIndividuals(sim.subpopulations.individuals);
}
9:11 late() {  // Last three years write everyone out to a separate VCF file named pX-T.vcf
               // where X is the subpop.id and T is the cycle number
  for (subpop in sim.subpopulations) {
    subpop.outputVCFSample(subpop.individualCount, replace=F, filePath="p" + subpop.id + "-" + sim.cycle + ".vcf");
  }
}

11 late() {
	sim.treeSeqOutput("SLiM.trees"); //  write out trees
}
```

Now we can run it in SLiM
```{r}
# run it via system
system(paste("slim -s 333 ", file, collapse = " "))
```

That writes out a lot of vcf files of genotypes of simulated individuals:
```{r}
dir(pattern = "p[0-9]+-[0-9]+\\.vcf")
```

It also writes out a file called `SLiM.trees`.

## Processing the trees

There are 11 years of simulation, so we want to get things for tree-seq years
0 to 10 from populations 1 and 2. We do that like this:
```{r}
# read the trees. Dump just captures the python code that was
# evaluated to get things in the py variable in the global env.
dump <- read_and_filter_trees(
  trees_path = "SLiM.trees",
  years_list = list(`1` = 0:10, `2` = 0:10)
)

```


Now that we have read and simplified that tree into the `py$ts` object,
we can turn it into tibbles of nodes and indivs easily:
```{r}
ni <- ts_nodes_and_inds(py$ts)
```

Now we get a vector of the founder nodes.  These are the ones
that existed at time 10 (counting backwards from 0) in TS time,
which is time 1 in SLiM time.
```{r}
founders <- founder_node_ids_and_pops(ni, 10)
```

Now we want to get the focal nodes.  These could be just the individuals
in the final three generations, but I want to plot pictures of the admixture
process in action, so I am going to make the focal individuals everyone
in SLiM year 2 and forward (so, TS time 9 and forward to 0), in both pops 1 and 2.
We designate the desired years and pops in a tibble.  The population labels
that come out of SLiM and are put into tskit at this point are base-1 indexed,
so they are named 1 and 2, not 0 and 1.
```{r}
focal_tibble <- tibble(
  pop = rep(c(1,2), each = 10),
  time = rep(0:9, 2)
)

focal_tibble
```

And we can use it in here like so:
```{r}
fnl <- get_focal_nodes(
  ni, 
  Focal = focal_tibble
)
```

Note that `fnl` has these focal nodes divided into different years, and the following
function cycles over those years and finds ancestors for each separately, because
otherwise some samples become the linked ancestors (in `link_ancestors` from tskit),
thus excluding the actual founders (this is why we couldn't use
[tspop](https://tspop.readthedocs.io/en/latest/tspop.html) for this directly---that
does not seem to handle multiple cohorts of samples)
which we absolutely need here.

Now we get the ancestral segments in all of them.
```{r}
segments_tib <- ancestral_segs(
  ts = py$ts,
  focal_nodes_list = fnl,
  founder_nodes = founders$nodes,
  nodes_tib = ni$nodes_tib,
  indiv_tib = ni$indiv_tib
)
```


## Plotting the _haploid_ ancestry segments

This is done with a simple function:
```{r, fig.height=10, fig.width=8}
plot_ancestry_of_genomes(segments_tib)
```


## Obtaining and plotting the ancestry tracts within each individual

There is a function for this:
```{r, message=FALSE}
indiv_segs <- indiv_ancestry_tracts(segments_tib)
```

And we can plot those like this:
```{r, fig.height=10, fig.width=8}
plot_ancestry_tracts(indiv_segs, chrom_ends = mykiss_chroms)
```

The left column is population 1 which is the WCT pop here. And the right column
is population 2 which is RBT.  

The colors denote the different ancestry combinations, which are coded in decimal
equivalents of trits or ternary numbers.  2 = 2 doses from p1; 4 = 1 dose from p1
and one from p2; 6 is two doses from p2.  This can extend to more than two
species.  

Perhaps more explanation is needed about the trits.  We envision the number of
doses from each population as a ternary number.  In a ternary number, each 
digit can be a 0, 1, or 2.  In the 1's column ($3^0$), the 0, 1 or 2, signify
just that---a 0, 1, or 2.  In the 3's column, the values are $0=0$, $1=3$ and 
$2=6$.  And so forth (just like binary, but base 3). Since we are counting doses
the sum of all the digits themselves must be two.  Here are some examples for
a three-species case:

- 002: two copies from species 1 --- $3^2 \times 0 + 3^1 \times 0 + 3^0 \times 2 = 2$
- 011: one copy from species 1 and one copy from species 2 --- $3^2 \times 0 + 3^1 \times 1 + 3^0 \times 1 = 4$
- 020: two copies from species 2 --- $3^2 \times 0 + 3^1 \times 2 + 3^0 \times 0 = 6$
- 200: two copies from species 3 --- $3^2 \times 2 + 3^1 \times 0 + 3^0 \times 0 = 18$

Coding things this way would make it easy to extend to even more species in the future,
if needed.

## Calculating admixture fractions from the ancestry tracts

This is easily done.
```{r}
Q_values <- admixture_fracts_from_ancestry_tracts(indiv_segs)
```


## Getting into the VCFs

The function `slim_vcf2tib()` reads a single VCF into a long format. It uses bcftools
via a pipe to make this pretty quick and painless.  It should be noted that a long format
for this is not space-efficient at all, but for the scale of data available for
the trout, we can deal with that, and it makes it pretty convenient for other things.


```{r}
p1_10_genos <- slim_vcf2tib("p1-10.vcf")
head(p1_10_genos, n = 10)
```

Note that you can pass that function `long = FALSE` and it will return a wide tibble
that might be useful.  (Though, in practice it only saves about 50% of the memory...).
```{r}
p1_10_wide <- slim_vcf2tib("p1-10.vcf", long = FALSE)

# look at a piece of that
head(p1_10_wide[, 1:9], n = 10)
```

In practice, most of the time we will be reading in multiple VCF files.  For that we 
have another function `slim_vcfs2tib()` that allows you to specify
the pops and years that you want to read. It also parses the file name and drops
some of the columns you don't need.  The years are specified in SLiM years
in a list named by the population.  For example, to get the slim years 9, 10, and 11 from
pops p1 and p2, we do this:
```{r}
sample_genos_long <- slim_vcfs2tib(Flist = list(p1 = 9:11, p2 = 9:11))
```

That is a big tibble with 7.5 million rows, but it does make it quick and easy
to calculate some things as demonstrated in the next section.

### A Note on Nomenclature

Throughout this package, we use `ped_id` to be the Pedigree ID _of the individual_.
SLiM's VCF has a `slimGenomePedigreeIDs` field.  These are the pedigree IDs of the
different haplotypes/gametes, _not_ of the individuals.  We don't really use those
much after we get the individual's pedigree IDs.  So, keep in mind...`ped_id` for us
is all about pedigree ID _of the individual_.  (Same with `ped_p1` and `ped_p2`---those
are the _individual_ pedigree IDs of the parents).

## Finding pairwise relationships amongst the sampled individuals

When we go to assess how well parentage can be done in these cases, we will be
curious to see how often aunts and uncles end up being called as parents, etc.
In order to make that sort of tally, we need to know the actual relationships of
all the sampled individuals.  We have the pedigree that connects them all, and we can
use that to find those relationships.  To do this we employ some functions from
the [CKMRpop](https://eriqande.github.io/CKMRpop/index.html) package that Eric
wrote.  Though we keep those wrapped in a function.  We can use it to find
the relatives amongst all the 6000 individuals in the last 3 generations
like this:
```{r}
pairwise_relats <- pairwise_relationships(
  IT = ni$indiv_tib,
  sample_years = 0:2,
  num_gen = 2
)
```

We can really quickly count up all the relationships here that we tallied in that
step:
```{r}
pairwise_relats %>%
  count(dom_relat, max_hit)
```

Key to these relationsihps:

- `A` is "avuncular", meaning aunt-niece or uncle-niece, etc.
- `FC` is first cousin
- `GP` is grandparent-grandchild
- `PO1 is parent-offspring
- `Si` is sibling

The `max_hit` column records how many ancestors are shared between these individuals
at this level of relationship.  So `Si` with `max_hit = 1` is a half-sibling, while
`Si` with `max_hit = 2` is a full-sibling.

Looking at this, it is clear that we will want to also use a modified SLiM script
in which the fish are not totally random mating---there are not enough full-siblings
in here, compared to half siblings.  (Hence in the MixedUpParents paper we also used
a non-Wright-Fisher model in SLiM which produced many more full-siblings).


## Simulating the diagnostic markers

We can simulate diagnostic markers in all of the individuals in the last three
years pretty easily.  The cyclone creek diagnostic markers are available as
package data in MixedUpSlimSims.  We will place them in
individuals of the last three years.
```{r}
sim_diag <- example_spp_diag_markers %>%
  cumulative_genome_position(mykiss_chroms) %>%
  rename(diag_spp_idx = diag_spp) %>%
  place_diagnostic_markers(IndivSegs = indiv_segs %>% filter(ind_time %in% 0:2))
```

## Writing out those two data sets


```{r}
set.seed(1)

subsamp <- sample(unique(sim_diag$ped_id), size = 2500, replace = FALSE)

sim_diag_ready <- sim_diag %>%
  filter(ped_id %in% subsamp) %>%
  rename(diag_spp = diag_spp_idx) %>%
  mutate(diag_spp = case_match(diag_spp, 1 ~ "WCT", 2 ~ "RBT", 3 ~ "YCT")) %>%
  rename(n = geno, indiv = ped_id) %>%
  mutate(indiv = as.character(indiv), pos = as.integer(pos), n = as.integer(n))

# the variable markers are a little more work because we have to get back to the
# original mykiss genome positions
cc_freqs3 <- cc_freqs2 %>%
  select(pos, chrom, cPOS)

sim_var_snps_ready <- sample_genos_long %>%
  filter(ped_id %in% subsamp) %>%
  select(pos, ped_id, haplo, allele)  %>%
  rename(cPOS = pos) %>%
  left_join(cc_freqs3, by = join_by(cPOS)) %>%
  select(chrom, pos, ped_id, allele) %>%
  group_by(chrom, pos, ped_id) %>%
  summarise(n = sum(as.integer(allele))) %>%
  ungroup() %>%
  arrange(ped_id, chrom, pos) %>%
  rename(indiv = ped_id) %>%
  mutate(indiv = as.character(indiv), pos = as.integer(pos))

```

And we also want to store some information about all the individuals---basically the metadata
and the pedigree, and the true Q-values.  That is all stored in the Q_values variable. And I also
want to keep the actual segments for each of the fish that we have the genotypes for.  
```{r}
Q_values

true_segs <- indiv_segs %>%
  filter(ped_id %in% subsamp)
```

If we wanted to write these two marker data sets out along with the "truth,"
we could do this:
```r
dir.create("results/004", showWarnings = FALSE, recursive = TRUE)
sim_results <- list(
  spp_diag = sim_diag_ready,
  var_snps = sim_var_snps_ready,
  metaQ = Q_values,
  true_segs = true_segs,
  cumul_genome = mykiss_chroms,
  alle_freqs = cc_freqs
)
write_rds(sim_results, file = "results/004/sim_results.rds", compress = "xz")
```


## Side note about development

As a side note.  For development, so that I still have the right
condaenv after building and relaoding the package in RStudio, I
have an uncommitted file called `R/onLoad.R` that has these contents:
```{.r}
# # global references to python modules (will be initialized in .onLoad)


.onLoad <- function(libname, pkgname) {
  reticulate::use_condaenv(condaenv = Sys.getenv("MUP_CONDA"))
}

```

And I have an environment variable in my .bashrc called MUP_CONDA
that points to the right spot.
